Noncoherent least squares estimators of carrier phase and amplitude

### 2012-08-02 other/confnoncoherent/paper.tex
more conf edits.
Added to paper

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK), and $M$-PSK.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normaly distributed.  However, the amplitude estimator is not consistent, and we show that it converges to a positive real number that is a function of the true carrier amplitude, the noise distribution, and the size of the constellation.  The results of Monte Carlo simulations are provided and these corroborate the theoretical results.
\end{abstract}
\begin{IEEEkeywords}
Noncoherent detection, differential phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

In passband communications systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation.  These effects must be compensated for at the reciever. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle, commonly refered to as $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M$ is the size of the constellation.  %We denote the set of symbols contained in the $M$-PSK constellation by $\calC$.

We assume that time offset estimation and matched filtering have already been performed at the reciever.  The recieved symbols then take the form,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i = 1, \dots, L,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from $y_1, \dots, y_L$.  If the transmitted symbols $s_1, \dots, s_L$ are known apriory at the reciever then the least squares estimator is typically used,
\[
\hat{a}_{\text{uc}} = \arg \min_{a \in \complex} \sum_{i = 1}^L \abs{ y_i - a s_i }^2  = \frac{1}{L} \sum_{i = 1}^L y_i s_i^*,
\]
where $\complex$ is the set of complex numbers and $x^*$ and $\abs{x}$ denotes the complex conjugate and complex magnitude of the complex number $x$.  Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983} call this the \emph{unmodulated carrier} estimator.  This estimator can be used if the tranmitter includes \emph{pilot} symbols, known to the reciever, i.e. with~\emph{coherent detection}.

Here we are interested in \emph{noncoherent detection}, when $s_1, \dots, s_L$ are not known at the reciever, and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009}.  A practical approach is the least squares estimator,
\[
\hat{a} = \arg\min_{a \in \complex} \min_{s_1, \dots, s_L \in \calC} \sum_{i = 1}^L \abs{ y_i - a s_i }^2,
\]
where $\calC$ is the set of symbols from the $M$-PSK constellation.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  As we will show, the estimator works well under less stringent assumptions.  Mackenthun~\cite{Mackenthun1994} described an algorithm to compute the least squares estimator $\hat{a}$ that requires only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.

@@

@@

\begin{corollary}\label{cor:ampmse}
The mean square error of the amplitude estimator $\hat{\rho}$ is 
\[
\expect (\rho_0 - \hat{\rho})^2 = \frac{B}{L} + \rho_0^2(1 - G(0))^2 + o(1)
\] 
where $o(1)$ goes to zero as $L \rightarrow \infty$.
\end{corollary}
\begin{IEEEproof} Write
\begin{align*}
\expect (\rho_0 - \hat{\rho})^2 &= \expect \big( \rho_0(1 - G(0)) + G(0)\rho_0 - \hat{\rho} \big)^2 \\
&= \expect \big( \rho_0(1 - G(0)) + \hat{m}_L \big)^2 \\
&= \rho_0^2(1 - G(0))^2 + \expect \hat{m}_L^2 + 2\rho_0(1 - G(0))\expect\hat{m}_L,
\end{align*}
and the corollary holds since $\expect \hat{m}_L^2 = B/L + o(1)$ and $\expect\hat{m}_L = o(1)$ from Theorem~\ref{thm:normality}.
\end{IEEEproof}

@@

A key assumption in Theorem~\ref{thm:consistency} is that the function $\abs{G(x)}$ is uniquely maximised at $x = 0$.

@@

>>>>

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normaly distributed.  However, the amplitude estimator is not consistent, it converges to a positive real number that is a function of the true carrier amplitude, the noise distribution, and the size of the constellation.  The results of Monte Carlo simulations are provided and these corroborate the theoretical results.
\end{abstract}
\begin{IEEEkeywords}
Noncoherent detection, differential phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

In passband communications systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation.  These effects must be compensated for at the reciever. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$. %and $M$ is the size of the constellation.  %We denote the set of symbols contained in the $M$-PSK constellation by $\calC$.

We assume that time offset estimation and matched filtering have already been performed at the reciever.  The recieved symbols are then,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i = 1, \dots, L,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from $y_1, \dots, y_L$.  If the transmitted symbols $s_1, \dots, s_L$ are known apriory at the reciever then the least squares estimator is typically used,
\[
\hat{a}_{\text{uc}} = \arg \min_{a \in \complex} \sum_{i = 1}^L \abs{ y_i - a s_i }^2  = \frac{1}{L} \sum_{i = 1}^L y_i s_i^*,
\]
where $\complex$ is the set of complex numbers and $x^*$ and $\abs{x}$ denotes the conjugate and magnitude of the complex number $x$.  Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983} call this the \emph{unmodulated carrier} estimator.  This estimator can be used if the tranmitter includes \emph{pilot} symbols, known to the reciever, i.e.~\emph{coherent detection}.

Here we are interested in \emph{noncoherent detection}, when $s_1, \dots, s_L$ are not known at the reciever, and must also be estimated.  This estimation problem has undergone extensive prior study and is often called \emph{multiple symbol differential detection}~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009}.  A practical approach is the least squares estimator,
\[
\hat{a} = \arg\min_{a \in \complex} \min_{s_1, \dots, s_L \in \calC} \sum_{i = 1}^L \abs{ y_i - a s_i }^2,
\]
where $\calC$ is the set of symbols from the $M$-PSK constellation.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  As we will show, the estimator can work well even when the noise is not Gaussian.  Mackenthun~\cite{Mackenthun1994} described an algorithm to compute the least squares estimator $\hat{a}$ that requires only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.

@@

The paper is organised in the following way.  Section~\ref{sec:circ-symm-compl} describes some properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} describes the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise sequence $\{w_i\}$, that $\langle\theta_0 - \hat{\theta}\rangle^2$ converges almost surely to zero and that $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle^2$ is asymptotically normally distributed.  However, the amplitude estimator $\hat{\rho_0}$ is not a consistent estimator of the true amplitude $\rho_0$.  The bias is small if the signal to noise ratio (SNR) is large, but is significant when the SNR is small.  In Section~\ref{sec:simulations} we present the results of simulations.  These simulations agree with the derived asymptotic properties.

@@

\begin{corollary}\label{cor:ampmse}
The mean square error of the amplitude estimator $\hat{\rho}$ is 
\[
\expect (\rho_0 - \hat{\rho})^2 = \frac{B}{L} + \rho_0^2(1 - G(0))^2 + o(1)
\] 
where $o(1)$ goes to zero as $L \rightarrow \infty$.
\end{corollary}
\begin{IEEEproof} Write
\begin{align*}
\expect (\rho_0 - \hat{\rho})^2 &= \expect \big( \rho_0(1 - G(0)) + G(0)\rho_0 - \hat{\rho} \big)^2 \\
&= \expect \big( \rho_0(1 - G(0)) - \hat{m}_L \big)^2 \\
&= \rho_0^2(1 - G(0))^2 + \expect \hat{m}_L^2 - 2\rho_0(1 - G(0))\expect\hat{m}_L,
\end{align*}
and the corollary holds since $\expect \hat{m}_L^2 = B/L + o(1)$ and $\expect\hat{m}_L = o(1)$ from Theorem~\ref{thm:normality}.
\end{IEEEproof}

@@

A key assumption in Theorem~\ref{thm:consistency} is that the function $\abs{G(x)}$ is uniquely maximised at $x = 0$.  The next Lemma describes a class of circularly symmetric distributions that satisfy this property.

@@

The theorems place conditions on the $\sfracpart{\hat{\theta} - \theta_0}$ rather than directly on $\hat{\theta} - \theta_0$.  This makes practical sense and BLERG.
